Tissue Enrichment Analysis (TEA)

### 2016-04-18 tests/hgf_benchmarking/doc/tea_main_text.tex
Minor changes
Plots are now saved as svg with a dpi=1200

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   % email={dangeles@caltech.edu}   % email address
]{\inits{DA}\fnm{David} \snm{Angeles-Albores}}
\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   % email={raymond@caltech.edu}   % email address
]{\inits{RYL}\fnm{Raymond Y} \snm{Lee}}
\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   % email={azurebrd@caltech.edu}   % email address
]{\inits{JC}\fnm{Juancarlos} \snm{Chan}}
\author[
   addressref={aff1},
   corref={aff1},
   email={pws@caltech.edu}
]{\inits{PWS}\fnm{Paul W} \snm{Sternberg}}

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
	RNA-seq and other high-throughput methods in biology have the ability to identify thousands of genes that are altered between conditions. These genes are often correlated in their biological characteristics or functions, but identifying these functions remains challenging. To interpret these long lists of genes, biologists need to abstract genes into fewer terms that are biologically relevant to form hypotheses about what is happening in the data. One such abstraction method relies on Gene Ontology (GO). GO provides a controlled set of hierarchically ordered terms~\cite{TheGeneOntologyConsortium2000a, TheGeneOntologyConsortium2015} that provide detailed information about the molecular, cellular or biochemical functions of any gene. For a given gene list, certain software programs can query whether a particular term is enriched~\cite{Mi2009, McLean2010, Huang2009}. One area of biological significance that GO does not include is anatomy. One way to address this shortcoming is to use a `tissue ontology' that provides a complete anatomical description for an organism (e.g `tissue', `organ' or `specific cell'), in this case for \emph{C.~elegans}~\cite{Lee2003}. The \emph{C.~elegans} database, WormBase~\cite{Howe2016}, maintains a curated list of gene expression data from GFP-reporters. Here we provide a new framework that analyzes a user-input list for enrichment of specific tissues. Tissues are physiologically relevant units with broad, relatively well-understood functionalities amenable to hypothesis formation. %As such, we believe that identification of tissues is likely to provide researchers with enough information to be able to form hypotheses about the physiological responses of an organism to a specified condition.

Another problem frequently associated with GO enrichment analysis is that it is often difficult to interpret due to the large number of terms associated with a given gene. There exist a number of GO analytic tools for use by the community but a shared complaint for many programs is the very large number of GO terms that are significantly associated with any given gene list. DAVID, a common tool for GO enrichment analysis, clusters enriched terms into broad categories~\cite{Huang2007}, whereas PANTHER~\cite{Mi2009, Mi2013} attempts to solve this issue by employing a manually reduced ontology, GOslim (H. Yu and P. Thomas, pers. comm. ).

@@

\section*{Results}
\subsection*{Generating a Useful Dictionary}
\subsubsection*{Reducing term redundancy through a similarity metric}
As a first step to generate our tissue enrichment software, we wished to select tissue terms that were reasonably well-annotated, yet specific enough to provide insight and not redundant with other terms. %We also wanted to avoid testing tissues at levels where annotation redundancy becomes problematic. For example, several left and right neurons have at least 25 annotated genes and we may want to include them for enrichment testing. 
For example, since many left/right neuronal pairs (which are sisters in the ontology) have almost identical annotations, with at most one or two gene differences between them%. We reasoned that when two tissues have almost identical annotations
, we cannot have statistical confidence in differentiating between them. As a result, testing these sister tissues provides no additional information compared with testing only the parent node to these sisters. %We refer to such sisters as `redundant'. 
To identify redundancy, we defined a similarity metric (see \emph{Methods} section and Figure~\ref{fig:simdiagram}) that can be used to identify sisters that have very high similarity between them; alternatively, redundant sisters could be identified if a single sister had a very high similarity score. We referred to these two scoring criteria as `avg' and `any' respectively.

@@

Additionally, we reasoned that, for any parent node, if all its daughters were selected for testing, there was no additional benefit to test the parent. %In other words, if all the daughter nodes are tested, there is little additional information to be gained by including the parent node. To address this issue 
We removed parent nodes from the analysis if all their daughter nodes passed the annotation threshold (see Figure~\ref{fig:trim_roots}). We called this a ceiling filter. Applying these three filters reduces multiple testing and verbosity.

@@

\subsection*{Tissue enrichment testing via a hypergeometric model}
Having built a static dictionary, we generated a Python script that implements a significance testing algorithm based on the hypergeometric model. Briefly, the hypergeometric model tests the probability of observing $n_i$ occurences of a tissue $i$ in a list of size $M$ if there are $m_i$ labels for that tissue in a dictionary of total size $N$ that are drawn without replacement. Mathematically, this is expressed as: 
\begin{eqnarray}\label{hypergeometric}
	\mathrm{P}(n_i | N, m_i, M) = \frac{\dbinom{m_i}{n_i} \dbinom{M-m_i}{N - n_i}}{\dbinom{N}{n_i}}\text{.}
\end{eqnarray}

@@

Users input a gene-list using any valid gene name for \emph{C.~elegans}. These names are processed into standard WormBase gene IDs (WBGene IDs). The program returns a table containing all the enriched terms and associated useful information %with significant tissues, their associated q-value, the expected number of hits for a list of that size, the observed number of hits and the enrichment fold change ($\text{observed hits} /\text{expected hits}$)
. Finally, the program can also return a bar chart of the enrichment fold change for the fifteen tissues with the lowest measured q-values. Our software is implemented in an easy to use GUI (see Figure~\ref{fig:GUIresults}). Anatomy terms are displayed in human-readable format followed by their unique ontology ID (WBbt ID).

\subsection*{Validation of the algorithm and parameter selection}
We wanted to select a dictionary that included enough terms to be specific beyond the largest \emph{C.~elegans} tissues, yet would minimize the number of spurious results and which had a good dynamic range in terms of enrichment fold-change. Larger tissues are correlated with better annotation, so increasing term specificity is associated with losses in statistical power.
%The phrase below doesn't seem useful
%To our knowledge there is no good method for assessing false-positive or false-negative results for annotations.
% Gold-standard data sets were annotated gene lists derived from the corresponding Expression Cluster data available in the WS252 version of WormBase.
To help us select an appropriate dictionary and validate our tool, we used a set of 30 gold standards based on microarray and RNA-seq literature which are believed to be enriched in specific tissues~\cite{Gaudet2004a, Spencer2011, Cinar2005, Watson2008a, Pauli2006, Portman2004, Fox2007, Smith2010}. These data sets are annotated gene lists derived from the corresponding Expression Cluster data in WormBase. Some of these studies have been used to annotate gene expression, and so they did not constitute an independent testing set. To correct this flaw, we built a clean dictionary that specifically excluded all annotation evidence that came from these studies.

As a first attempt to select a dictionary, we generated all possible combinations of dictionaries with minimal annotations of 10, 25, 50 and 100 genes and similarity cutoffs of 0.9, 0.95 and 1, using `avg' or `any' similarity thresholding methods (see Table~\ref{tab:DictionarySpecs}). The number of tissues was inversely correlated to the minimum annotation cutoff, and was largely insensitive to the similarity threshold in the range we explored. Next, we analyzed all 30 datasets using each dictionary. Because of the large number of results, instead of analyzing each set of terms individually, we measured the average q-value for significantly enriched terms in each dataset. When we analyzed the distribution of significant q-values for the dictionaries, we found that the similarity threshold mattered relatively little for any dictionary. We also noticed that the `any' thresholding method resulted in tighter histograms with a mode closer to 0. For this reason, we chose the `any' method for dictionary generation. The average q-value increased with decreasing annotation cut-off (see Figure~\ref{fig:qvals}), which reflects the decreasing statistical power associated with fewer annotations per term, but we remained agnostic as to how significant is the trade-off between power and term specificity. Based on these observations, we ruled out the dictionary with the 100 gene annotation cut-off: it had the fewest terms and its q-values were not low enough in our opinion to compensate the trade-off in specificity.

%dictionaries that work well
To select between dictionaries generated between 50, 33 and 25 annotation cut-offs, and also to ensure the terms that are selected as enriched by our algorithm are reasonable, we looked in detail at the enrichment analysis results.
%Work well
% TODO add WBbt codes!
Most results were comparable and expected. For some sets, all dictionaries performed well. For example, in our `all neuron enriched sets'~\cite{Spencer2011, Watson2008a} all terms were neuron-related %including mechanosensory neurons, thermosensitive neurons, interneurons, ganglions and male rays 
regardless of the dictionary used (see Table~\ref{tab:interagree}). On the other hand, for a set enriched for germline precursor expression in the embryo~\cite{Spencer2011}, the 50 cutoff dictionary was only able to identify `oocyte WBbt:006797' which is not a germline precursor, although it is germline related; whereas the two smaller dictionaries singled out actual germline precursor cells---at the 33 cutoff, our tool identified the larval germline precursor cells `Z2' and `Z3' as enriched, and at the 25 gene cutoff the embryonic germline precursor terms `Psub4',`Psub3' and `Psub2' were identified in addition to `Z2' and `Z3'. 
%Differences between dictionaries
We also queried an intestine precursor set~\cite{Spencer2011}. Notably, this gene set yielded no enrichment when using the 25 cutoff dictionary, nor when using the 50 cutoff dictionary. However, the 33 cutoff dictionary identified the E lineage, which is the intestinal precursor lineage in \emph{C.~elegans}, as enriched. Both of these results capture specific aspects of \emph{C.~elegans} that are well known to developmental biologists.

@@

\subsection*{A brief example}
We applied our tool to the RNA-seq datasets developed by Engelmann et al.~\cite{Engelmann2011} to gain further understanding of their underlying biology. Engelmann et al. exposed young adult worms to 5 different pathogenic bacteria or fungi for 24 hours, after which mRNA was extracted from the worms for sequencing. We ran TEA on the genes Engelmann \emph{et al} identified as up- or down-regulated. Initially we noticed that genes that are down-regulated tend to be twice better annotated on average than genes that were up-regulated, suggesting that our understanding of the worm immune system is scarce, in spite of important advances made over the last decade. Up-regulated tissues, when detected, almost always included the hypodermis and excretory duct. Three out of the five samples showed enrichment of neuronal tissues or neuronal precursor tissues amongst the down-regulated genes. As an independent verification, we also performed GO analysis using PANTHER on the down-regulated genes for \emph{D.~coniospora}. These results showed enrichment in terms associated with neurons (see Figure~\ref{fig:Dcon}). A possible explanation for this neuronal association might be that the infected worms are sick and the neurons are beginning to shut down; an alternative hypothesis would be that the worm is down-regulating specific neuronal pathways as a behavioral response against the pathogen. Indeed, several studies~\cite{Meisel2014, Zhang2005} have provided evidence that \emph{C.~elegans} uses chemosensory neurons to identify pathogens. %One bacterium did not ellicit the same pattern of down-regulation of neuronal-associated genes. \emph{E.~faecalis} showed increased expression of genes associated with neuronal tissues, hinting that \emph{E.~faecalis} may have a different pathogenic profile. 
Our results highlight the involvement of various \emph{C.~elegans} neuronal tissues in pathogen defense.

@@

%Other tools
Our tool relies on an annotation dictionary that is continuously updated, does not require retraining and does not require ranked genes. To our knowledge, this is the first tool that tests tissue enrichment  in \emph{C.~elegans} via the hypergeometric method, but similar projects exist for humans and zebrafish~\cite{Lee2013, Prykhozhij2013}, highlighting the relevance of our tool for high-dimensionality biology. Chikina \emph{et al}~\cite{Chikina2009} have previously reported a tissue enrichment model for \emph{C.~elegans }based on a Support Vector Machine classifier that has been trained on microarray studies. SVMs are powerful tools, but they require continuous retraining as more tissue expression data becomes available. Moreover, classifiers require that data be rank-ordered by some metric, something which is not possible for certain studies. Moreover, this tissue enrichment tool generally provides users with enrichment results for 6 large tissues. In contrast, our tool routinely tests a much larger number of terms, and we have shown it can even accurately identify enrichment of embryonic precursor lineages for select data sets.

@@

We have endeavoured to benchmark our tool well. However, our analysis could not identify %identify good negative controls
 spurious term enrichment.
% Even for our set of positive controls, the statistical analysis sometimes throws out unexpected results. For example, the embryonic germline precursor gene set had the term `AB', which refers to the anterior cell in the 2-cell stage \emph{C.~elegans} embryo, as the most enriched term in the dictionaries with cut off of 25 and 33. Is this an error, or does this hint at new biology? 
 Although we were unable to determine false-positive and false-negative rates, we do not believe this should deter scientists from using our tool. Rather, we encourage researchers to use our tool as a guide, integrating evidence from multiple sources to inform the most likely hypotheses. As with any other tool based on statistical sampling, our analysis is most vulnerable to bias in the data set. For example, expression reports are negatively biased against germline expression because of the difficulties associated with expressing DNA in this tissue~\cite{Kelly1997}. As time passes, we are certain the accuracy and power of this tool will improve thanks to the efforts of the combined worm community; indeed, without the community reports of tissue expression in the first place, this tool would not be possible.

@@

\texttt{pip install tissue\_enrichment\_tool}

@@

% Figures describing algorithm

\begin{figure}[h!]
  \begin{subfigure}[b]{0.3\textwidth}
    \includegraphics[width=\textwidth]{figures/RedundancyTrimmingOntogeny.png}
   	 	\caption{}
    \label{fig:simdiagram}
  \end{subfigure}
  %
  \begin{subfigure}[b]{0.3\textwidth}
    \includegraphics[width=\textwidth]{figures/FloorTrimmingOntogeny.png}
	      \caption{}
    \label{fig:trim_ends}
  \end{subfigure}
  \begin{subfigure}[b]{0.3\textwidth}
    \includegraphics[width=\textwidth]{figures/CeilingTrimmingOntogeny.png}
    	\caption{}
    \label{fig:trim_roots}
  \end{subfigure}

  \captionsetup{width= 0.95\textwidth}
  \caption{
  	\csentence{Schematic representation of trimming filters for an acyclical ontology.}
    \csentence{\textbf{a)}} The parent node (green) contains at least as many annotations as the union of the two sisters. These two sisters share annotations extensively. Therefore they are too similar and should be removed. 	\csentence{\textbf{b)}} Nodes with less than a threshold number of genes are trimmed (red) and discarded from the dictionary. Here, the example threshold is 25 genes. Nodes $\epsilon, \zeta, \eta$, shown in red will be removed.
	\csentence{\textbf{c)}} We trim parent nodes recursively, starting from the root, if all their daughter nodes have more than the threshold number of annotations. Nodes in grey ($\epsilon, \zeta, \eta$) were removed in the previous step. Nodes $\alpha, \beta$ shown in red will be trimmed because each one has a complete daughter set. Only nodes $\gamma$ and $\delta$ will be used to generate the static dictionary. 
  }
\end{figure}

%gui
\begin{figure}
	\centering
    \includegraphics[width=0.95\textwidth]{figures/guiresults.png}
  	\captionsetup{width= 0.95\textwidth}
 	\caption{\csentence{Screenshot of results from the web GUI.}
  After inputting a gene-list, the user is provided with the results. 
  An HTML table is output with hyperlinks to the ontology terms. A 
  publication-ready graph is provided below, which can be saved by 
  dragging to the desktop. The graph and the table show anatomy terms in
  human-readable format, followed by their unique WBbt ID. Finally, lists 
  of the genes used and discarded for the analysis are also presented.
  }
  \label{fig:GUIresults}
\end{figure}

% statistical testing and validation Figures
%q plot
\begin{figure}[h!]
	\includegraphics[width=\textwidth]{avgQKDE_method=any.pdf}
  \captionsetup{width= 0.95\textwidth}
  \caption{\csentence{Kernel density estimates for 30 gold standard datasets.}
      We ran TEA on 30 datasets we believed to be enriched in particular tissues and pooled all the results to observe the distribution of q-values. The mode of the distribution for dictionaries with annotation cut-offs of 100 and 50 genes are very similar; however, when the cut-off is lowered to 25 genes, the mode of the distribution shifts to the left, potentially signalling a decrease in measurement power.}
	  \label{fig:qvals}
\end{figure}

%pan neuronal inter dict agreement
% \begin{figure}
%   \begin{subfigure}[b]{0.95\textwidth}
%     \includegraphics[width=0.95\textwidth]{WBPaper00031532_Larva_Pan_Neuronal_Enriched_WBbt_0003679_1603_33cutoff.png}
%     \caption{}
%     \label{fig:PanNeuronalWatson33}
%   \end{subfigure}
%   %
%   \begin{subfigure}[b]{0.95\textwidth}
%     \includegraphics[width=0.95\textwidth]{WBPaper00031532_Larva_Pan_Neuronal_Enriched_WBbt_0003679_1603_50cutoff}
%     \caption{}
%     \label{fig:PanNeuronalWatson50}
%   \end{subfigure}
%   \captionsetup{width= 0.95\textwidth}
%   \caption{\csentence{Comparison of Enrichment Results for two dictionaries for a pan-neuronal enriched gene set from Watson, 2008~\cite{Watson2008a}.}
% 	Bar labels are anatomy terms in human-readable format, followed by their unique WBbt ID.
%    \textbf{a)} Dictionary with cutoff: 33; threshold: 0.95; method: `any'
%    \textbf{b)} Dictionary with cutoff: 50; threshold: 0.95; method: `any'
%    }
%    \label{fig:interagree}
% \end{figure}

>>>>

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   % email={dangeles@caltech.edu}   % email address
]{\inits{DA}\fnm{David} \snm{Angeles-Albores}}
\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   % email={raymond@caltech.edu}   % email address
]{\inits{RYL}\fnm{Raymond YN} \snm{Lee}}
\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   % email={azurebrd@caltech.edu}   % email address
]{\inits{JC}\fnm{Juancarlos} \snm{Chan}}
\author[
   addressref={aff1},
   corref={aff1},
   email={pws@caltech.edu}
]{\inits{PWS}\fnm{Paul W} \snm{Sternberg}}

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
	RNA-seq and other high-throughput methods in biology have the ability to identify thousands of genes that are altered between conditions. These genes are often correlated in their biological characteristics or functions, but identifying these functions remains challenging. To interpret these long lists of genes, biologists need to abstract genes into concepts that are biologically relevant to form hypotheses about what is happening in the system. One such abstraction method relies on Gene Ontology (GO). GO provides a controlled set of hierarchically ordered terms~\cite{TheGeneOntologyConsortium2000a, TheGeneOntologyConsortium2015} that provide detailed descriptions about the molecular, cellular or biochemical functions of any gene. For a given gene list, certain software programs can query whether a particular term is enriched~\cite{Mi2009, McLean2010, Huang2009}. One area of biological significance that GO does not include is anatomy. One way to address this shortcoming is to use a `tissue ontology' that provides a complete anatomical description for an organism (e.g `tissue', `organ' or `specific cell'), in this case for \emph{C.~elegans}~\cite{Lee2003}. The \emph{C.~elegans} database, WormBase~\cite{Howe2016}, maintains a curated list of gene expression data from the literature. Here we provide a new framework that analyzes a user-input list for enrichment of specific cells and tissues. They are physiologically relevant units with broad, relatively well-understood functionalities amenable to hypothesis formation. %As such, we believe that identification of tissues is likely to provide researchers with enough information to be able to form hypotheses about the physiological responses of an organism to a specified condition.

Another problem frequently associated with GO enrichment analysis is that it is often difficult to interpret due to the large number of terms associated with a given gene. There exist a number of GO analytic tools for use by the community but a shared complaint for many programs is the very large number of GO terms that are significantly associated with any given gene list. DAVID, a common tool for GO enrichment analysis, clusters enriched terms into broad categories~\cite{Huang2007}, whereas PANTHER~\cite{Mi2009, Mi2013} attempts to solve this issue by employing a manually reduced ontology, GOslim (H. Yu and P. Thomas, pers. comm.).

@@

\section*{Results}
\subsection*{Generating a Gene-Tissue Dictionary}
\subsubsection*{Reducing term redundancy through a similarity metric}
As a first step to generate our tissue enrichment software, we wished to select tissue terms that were reasonably well-annotated, yet specific enough to provide insight and not redundant with other terms. %We also wanted to avoid testing tissues at levels where annotation redundancy becomes problematic. For example, several left and right neurons have at least 25 annotated genes and we may want to include them for enrichment testing. 
For example, nematodes have a number of neurons that are placed symmetrically along the left/right body axis, and are functionally similar. These left/right neuronal pairs (which are sisters in the ontology) have almost identical annotations, with at most one or two gene differences between them%. We reasoned that when two tissues have almost identical annotations
, and therefore we cannot have statistical confidence in differentiating between them. As a result, testing these sister tissues provides no additional information compared with testing only the parent node to these sisters. %We refer to such sisters as `redundant'. 
To identify redundancy, we defined a similarity metric (see \emph{Methods} section and Figure~\ref{fig:simdiagram}) that can be used to identify sisters that have very high similarity between them; alternatively, redundant sisters could be identified if a single sister had a very high similarity score. We referred to these two scoring criteria as `avg' and `any' respectively.

@@

Additionally, we reasoned that, for any parent node, if all its daughters were selected for testing, there was no additional benefit to test the parent. %In other words, if all the daughter nodes are tested, there is little additional information to be gained by including the parent node. To address this issue 
We removed parent nodes from the analysis if all their daughter nodes passed the annotation threshold (see Figure~\ref{fig:trim_roots}). We called this a ceiling filter. Applying these three filters reduced the number of ontology terms by an order of magnitude.

@@

\subsection*{Tissue enrichment testing via a hypergeometric model}
Having built a static dictionary, we generated a Python script that implements a significance testing algorithm based on the hypergeometric model. Briefly, the hypergeometric model tests the probability of observing $n_i$ occurences of a tissue $i$ in a list of size $M$ if there are $m_i$ labels for that tissue in a dictionary of total size $N$ that are drawn without replacement. Mathematically, this is expressed as:

\begin{eqnarray}\label{hypergeometric}
	\mathrm{P}(n_i | N, m_i, M) = \frac{\dbinom{m_i}{n_i} \dbinom{M-m_i}{N - n_i}}{\dbinom{N}{n_i}}\text{.}
\end{eqnarray}

@@

Users input a gene list using any valid gene name for \emph{C.~elegans}. These names are processed into standard WormBase gene IDs (WBGene IDs). The program returns a table containing all the enriched terms and associated useful information%with significant tissues, their associated q-value, the expected number of hits for a list of that size, the observed number of hits and the enrichment fold change ($\text{observed hits} /\text{expected hits}$)
. Finally, the program can also return a bar chart of the enrichment fold change for the fifteen tissues with the lowest measured q-values. The bars in the graph are sorted in ascending order of q-value and then in descending order of fold-change. Bars are colored for ease of viewing, and color does not convey information. Our software is implemented in an easy to use GUI (see Figure~\ref{fig:GUIresults}). Anatomy terms are displayed in human-readable format followed by their unique ontology ID (WBbt ID). In summary, each time the ontology annotations are updated, a new trimmed ontology is generated using our filters; in parallel, users can submit their gene lists through WormBase for testing, with results output in a number of formats (see Figure~\ref{fig:workflow}).

\subsection*{Validation of the algorithm and optimizing parameter selection}
We wanted to select a dictionary that included enough terms to be specific beyond the largest \emph{C.~elegans} tissues, yet would minimize the number of spurious results and which had a good dynamic range in terms of enrichment fold-change. Larger tissues are correlated with better annotation, so increasing term specificity is associated with losses in statistical power.
%The phrase below doesn't seem useful
%To our knowledge there is no good method for assessing false-positive or false-negative results for annotations.
% Gold-standard data sets were annotated gene lists derived from the corresponding Expression Cluster data available in the WS252 version of WormBase.
To help us select an appropriate dictionary and validate our tool, we used a set of 30 gold standards based on microarray and RNA-seq literature which are believed to be enriched in specific tissues~\cite{Gaudet2004a, Spencer2011, Cinar2005, Watson2008a, Pauli2006, Portman2004, Fox2007, Smith2010}. These data sets are annotated gene lists derived from the corresponding Expression Cluster data in WormBase. Some of these studies have been used to annotate gene expression, and so they did not constitute an independent testing set. To correct this flaw, we built a clean dictionary that specifically excluded all annotation evidence that came from these studies.

As a first attempt to select a dictionary, we generated all possible combinations of dictionaries with minimal annotations of 10, 25, 50 and 100 genes and similarity cutoffs of 0.9, 0.95 and 1, using `avg' or `any' similarity thresholding methods (see Table~\ref{tab:DictionarySpecs}). The number of tissues was inversely correlated to the minimum annotation cutoff, and was largely insensitive to the similarity threshold in the range we explored. Next, we analyzed all 30 datasets using each dictionary. Because of the large number of results, instead of analyzing each set of terms individually, we measured the average q-value for significantly enriched terms in each dataset without regard for the perceived accuracy of the terms that tested significant. We found that the similarity threshold mattered relatively little for any dictionary. We also noticed that the `any' thresholding method resulted in tighter histograms with a mode closer to 0. For this reason, we chose the `any' method for dictionary generation. The average q-value increased with decreasing annotation cut-off (see Figure~\ref{fig:qvals}), which reflects the decreasing statistical power associated with fewer annotations per term, but we remained agnostic as to how significant is the trade-off between power and term specificity. Based on these observations, we ruled out the dictionary with the 100 gene annotation cut-off: it had the fewest terms and its q-values were not low enough in our opinion to compensate the trade-off in specificity.

%dictionaries that work well
To select between dictionaries generated between 50, 33 and 25 annotation cut-offs, and also to ensure the terms that are selected as enriched by our algorithm are reasonable, we looked in detail at the enrichment analysis results.
%Work well
% TODO add WBbt codes!
Most results were comparable and expected. For some sets, all dictionaries performed well. For example, in our `all neuron enriched sets'~\cite{Spencer2011, Watson2008a} all terms were neuron-related %including mechanosensory neurons, thermosensitive neurons, interneurons, ganglions and male rays 
regardless of the dictionary used (see Table~\ref{tab:interagree}). On the other hand, for a set enriched for germline precursor expression in the embryo~\cite{Spencer2011}, the 50 cutoff dictionary was only able to identify `oocyte WBbt:006797' which is not a germline precursor, although it is germline related; whereas the two smaller dictionaries singled out actual germline precursor cells---at the 33 cutoff, our tool identified the larval germline precursor cells `Z2' and `Z3' as enriched, and at the 25 gene cutoff the embryonic germline precursor terms `P$_4$',`P$_3$' and `P$_2$' were identified in addition to `Z2' and `Z3'. 
%Differences between dictionaries
We also queried an intestine precursor set~\cite{Spencer2011}. Notably, this gene set yielded no enrichment when using the 25 cutoff dictionary, nor when using the 50 cutoff dictionary. However, the 33 cutoff dictionary identified the E lineage, which is the intestinal precursor lineage in \emph{C.~elegans}, as enriched. Both of these results capture specific aspects of \emph{C.~elegans} that are well known to developmental biologists.

@@

\section*{Applying the tool}
We applied our tool to the RNA-seq datasets developed by Engelmann et al.~\cite{Engelmann2011} to gain further understanding of their underlying biology. Engelmann et al. exposed young adult worms to 5 different pathogenic bacteria or fungi for 24 hours, after which mRNA was extracted from the worms for sequencing. We ran TEA on the genes Engelmann \emph{et al} identified as up- or down-regulated. Initially we noticed that genes that are down-regulated tend to be twice better annotated on average than genes that were up-regulated, suggesting that our understanding of the worm immune system is scarce, in spite of important advances made over the last decade. Up-regulated tissues, when detected, almost always included the hypodermis and excretory duct. Three out of the five samples showed enrichment of neuronal tissues or neuronal precursor tissues amongst the down-regulated genes. As an independent verification, we also performed GO analysis using PANTHER on the down-regulated genes for \emph{D.~coniospora}. These results showed enrichment in terms associated with neurons (see Figure~\ref{fig:Dcon}). A possible explanation for this neuronal association might be that the infected worms are sick and the neurons are beginning to shut down; an alternative hypothesis would be that the worm is down-regulating specific neuronal pathways as a behavioral response against the pathogen. Indeed, several studies~\cite{Meisel2014, Zhang2005} have provided evidence that \emph{C.~elegans} uses chemosensory neurons to identify pathogens. %One bacterium did not ellicit the same pattern of down-regulation of neuronal-associated genes. \emph{E.~faecalis} showed increased expression of genes associated with neuronal tissues, hinting that \emph{E.~faecalis} may have a different pathogenic profile. 
Our results highlight the involvement of various \emph{C.~elegans} neuronal tissues in pathogen defense.

@@

%Other tools
Our tool relies on an annotation dictionary that is continuously updated primarily with data from single gene qualitative analyses, does not require retraining and does not require ranked genes. To our knowledge, this is the first tool that tests tissue enrichment  in \emph{C.~elegans} via the hypergeometric method, but similar projects exist for humans and zebrafish~\cite{Lee2013, Prykhozhij2013}, highlighting the relevance of our tool for high-dimensionality biology. Chikina \emph{et al}~\cite{Chikina2009} have previously reported a tissue enrichment model for \emph{C.~elegans }based on a Support Vector Machine classifier that has been trained on microarray studies. SVMs are powerful tools, but they require continuous retraining as more tissue expression data becomes available. Moreover, classifiers require that data be rank-ordered by some metric, something which is not possible for certain studies. Moreover, this tissue enrichment tool generally provides users with enrichment results for 6 large tissues. In contrast, our tool routinely tests a much larger number of terms, and we have shown it can even accurately identify enrichment of embryonic precursor lineages for select data sets.

@@

We have endeavoured to benchmark our tool well. However, our analysis could not identify %identify good negative controls
 spurious term enrichment.
% Even for our set of positive controls, the statistical analysis sometimes throws out unexpected results. For example, the embryonic germline precursor gene set had the term `AB', which refers to the anterior cell in the 2-cell stage \emph{C.~elegans} embryo, as the most enriched term in the dictionaries with cut off of 25 and 33. Is this an error, or does this hint at new biology? 
 Although we were unable to determine false-positive and false-negative rates, we do not believe this should deter scientists from using our tool. Rather, we encourage researchers to use our tool as a guide, integrating evidence from multiple sources to inform the most likely hypotheses. As with any other tool based on statistical sampling, our analysis is most vulnerable to bias in the data set. For example, expression reports are negatively biased against germline expression because of the difficulties associated with expressing DNA in this tissue~\cite{Kelly1997}. As time passes, we are certain the accuracy and power of this tool will improve thanks to the continuing efforts of the  worm research community; indeed, without the community reports of tissue expression in the first place, this tool would not be possible.

@@

\texttt{pip install tissue\_enrichment\_analysis}

@@

% Figures describing algorithm
\begin{figure}[h!]
  \begin{subfigure}[b]{0.3\textwidth}
    \includegraphics[width=\textwidth]{figures/RedundancyTrimmingOntogeny.png}
   	 	\caption{}
    \label{fig:simdiagram}
  \end{subfigure}
  %
  \begin{subfigure}[b]{0.3\textwidth}
    \includegraphics[width=\textwidth]{figures/FloorTrimmingOntogeny.png}
	      \caption{}
    \label{fig:trim_ends}
  \end{subfigure}
  \begin{subfigure}[b]{0.3\textwidth}
    \includegraphics[width=\textwidth]{figures/CeilingTrimmingOntogeny.png}
    	\caption{}
    \label{fig:trim_roots}
  \end{subfigure}

  \captionsetup{width= 0.95\textwidth}
  \caption{
  	\csentence{Schematic representation of trimming filters for an acyclical ontology.}
    \csentence{\textbf{a)}} The parent node (green) contains at least as many annotations as the union of the two sisters. These two sisters share annotations extensively, as expressed by the overlap in the Venn diagram, so they qualify for removal. 	
	\csentence{\textbf{b)}} Nodes with less than a threshold number of genes are trimmed (red) and discarded from the dictionary. Here, the example threshold is 25 genes. Nodes $\epsilon, \zeta, \eta$, shown in red are removed.
	\csentence{\textbf{c)}} Parent nodes are removed recursively, starting from the root, if all their daughter nodes have more than the threshold number of annotations. Nodes in grey ($\epsilon, \zeta, \eta$) were removed in the previous step. Nodes $\alpha, \beta$ shown in red are trimmed because each one has a complete daughter set. Only nodes $\gamma$ and $\delta$ will be used to generate the static dictionary. 
  }
\end{figure}

%gui
\begin{figure}
	\centering
    \includegraphics[width=0.95\textwidth]{figures/guiresults.png}
  	\captionsetup{width= 0.95\textwidth}
 	\caption{
	\csentence{Screenshot of results from the web GUI.} 
	After inputting a gene-list, the user is provided with the results. An HTML table is output with hyperlinks to the ontology terms. A publication-ready graph is provided below, which can be saved by  dragging to the desktop. The graph is colored for better visualization; color is not intended to convey information. The graph and the table show anatomy terms in  human-readable format, followed by their unique WBbt ID. Finally, lists  of the genes used and discarded for the analysis are also presented.
  }
  \label{fig:GUIresults}
\end{figure}

% workflow 
\begin{figure}
	\centering
	\includegraphics[width=0.95\textwidth]{figures/workflow.pdf}
	\captionsetup{width= 0.95\textwidth}
	\caption{
	\csentence{TEA Workflow.}
	The complete ontology is annotated continuously by WormBase curators. After each update, the ontology is processed to remove uninformative terms. The trimmed ontology is used for statistical testing. 
	Users can select a gene list and input it into our tool using our WormBase portal. The gene list is tested for enrichment using the trimmed ontology, and results are output in tabular and graphic formats for analysis. 
	}
	\label{fig:workflow}
\end{figure}

% statistical testing and validation Figures
%q plot
\begin{figure}[h!]
	\includegraphics[width=\textwidth]{avgQKDE_method=any.pdf}
  \captionsetup{width= 0.95\textwidth}
  \caption{
  \csentence{Kernel density estimates for 30 gold standard datasets.}
      We ran TEA on 30 datasets we believed to be enriched in particular tissues and pooled all the results to observe the distribution of q-values. The mode of the distribution for dictionaries with annotation cut-offs of 100 and 50 genes are very similar; however, when the cut-off is lowered to 25 genes, the mode of the distribution shifts to the left, potentially signalling a decrease in measurement power.
	  }
	  \label{fig:qvals}
\end{figure}

%gaba
\begin{figure}
  \includegraphics[width=0.95\textwidth]{GABAcomparison3.pdf}
  \captionsetup{width= 0.95\textwidth}
  \caption{
  \csentence{Independently derived gene sets show similar results 	when tested with the same dictionary.}
  \textbf{Set 1)} GABAergic gene set from Watson~\cite{Watson2008a}.
  \textbf{Set 2)} GABAergic gene set from Spencer~\cite{Spencer2011}.
  Arrowheads highlight identical terms between both analyses. All terms refer to neurons or neuronal tissues and are GABA-associated. Dictionary with cutoff: 33; threshold: 0.95; method: `any'.
   }
  \label{fig:intragree}
\end{figure}
